Visual narratives of the Covid-19 pandemic

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
phrasing
Frequency domain and nyt sparklines charts

@@

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro.png}
	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
	\label{fig:choro1}
\end{figure*}

\begin{figure}
\includegraphics[width=.99\linewidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{figure}

\begin{figure}
\includegraphics[width=.99\linewidth]{nyt-covid-scaled-choro}
\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
\end{figure}

@@

% Is it worth discussing e.g. https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem? When everything is always extremely high risk we just tend to tune things out. I don't want to distract from the overview, though...

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time. This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

@@

<<freq-domain-plot, include = F, eval = F>>=
show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")

n <- 50000
tmp <- cvstate %>%
  filter(state %in% show_states) %>%
  select(state, date, cases) %>%
  mutate(mod_num = cases %% n,
         whole_num = floor(cases/n)) %>%
  group_by(state, whole_num) %>%
  mutate(whole_num_days = n()) %>%
  group_by(state) %>%
  arrange(date)

tmp2 <- bind_rows(
  # Day before first case detected:
  filter(tmp, whole_num == 0) %>%
    group_by(state) %>%
    summarize(date = min(date) - lubridate::days(1), y = 0, whole_num = 0, 
              whole_num_days = unique(whole_num_days), mod_num = 0) %>%
    unique() %>%
    ungroup(),
  # Date before the count rolls over
  tmp %>%
    group_by(state, whole_num) %>%
    filter(date == max(date)) %>%
    group_by(state) %>%
    mutate(y = ifelse(whole_num == max(whole_num), mod_num, n),
           whole_num_days = ifelse(whole_num == max(whole_num), whole_num_days/(y/10000), whole_num_days)),
  # Date after the count rolls over
  tmp %>%
    group_by(state, whole_num) %>%
    filter(date == min(date)) %>%
    mutate(y = 0, date = date - lubridate::days(1))
  ) %>%
  group_by(state, whole_num) %>%
  mutate(whole_num_days = max(whole_num_days))

ggplot(data = tmp2, aes(x = date, ymin = 0, ymax = y, 
                        colour = whole_num_days, 
                        fill = whole_num_days, 
                        group = whole_num)) + 
  geom_ribbon() + 
  facet_grid(state~.) + 
  scale_x_date(date_breaks = "month", date_labels = "%B") +
  scale_y_continuous(breaks = c(n/2, n), labels = paste0(c(n/2000, n/1000), "k"))  + 
  scale_colour_gradientn(guide = 'none', paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
    trans = "log10",
    colours = darkblueseqpal
    ) + 
  scale_fill_gradientn(paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
    trans = "log10",
    colours = blueseqpal) + 
  my_theme + 
  theme(axis.title = element_blank(), 
        legend.background = element_rect(fill = "transparent", colour = NA), 
        legend.position = c(0, .66), legend.justification = c(0, 1), legend.direction = "horizontal")
@

@@

Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically novice viewers. Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case. Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.

>>>>

\usepackage{subcaption}

@@

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro}
	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
	\label{fig:choro1}
\end{figure*}

\begin{figure}
\centering
\begin{subfigure}[t]{.45\textwidth}
\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{subfigure}\hfill
\begin{subfigure}[t]{.45\textwidth}
\includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
\end{subfigure}
\end{figure}

@@

% Is it worth discussing e.g. https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem? When everything is always extremely high risk we just tend to tune things out. I don't want to distract from the overview, though... https://twitter.com/rypan/status/1282020934044401664 is a good example

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis. This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

@@

However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position. We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

\begin{figure}
\centering
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
\caption{Sparklines style heatmaps embedded in a table of coronavirus data by state. Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar. Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }

\label{fig:sparklines-heatmap-nyt}
\end{figure}

One unique time-series plot created a representation of the case counts in the frequency domain: instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle. This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}. Interestingly, this frequency representation can also be easily transferred into an audio domain, providing access for visually impaired users as well as a multimodal representation of the data for sighted users.

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss. Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
\label{fig:freq-domain}
\end{figure}

@@

Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically unsophisticated viewers. Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case. Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.
